### 2019-07-06 2-Chapters/3-Chapter/chapter3.tex
Small modifs
J'utilise moins "our ..."

% \chapter{SMPyBandits: a state-of-the-art Python library to simulate MAB problems}
\chapter{SMPyBandits: our Python library to simulate MAB problems}
\label{chapter:3}

@@

\paragraph{References.}
%
The code for SMPyBandits is hosted at \texttt{\href{https://GitHub.com/SMPyBandits/SMPyBandits/}{GitHub.com/SMPyBandits/SMPyBandits}}, its documentation is at \texttt{\href{https://SMPyBandits.GitHub.io/}{SMPyBandits.GitHub.io}}, and all the library is freely publicly released under the MIT open-source License.
This chapter is based on our article \cite{SMPyBanditsJMLR}.

@@

We chose to present here our Python library SMPyBandits in this chapter rather than in the Appendix, because all the numerical experiments on single-player multi-arm bandits in Chapter~\ref{chapter:25}, and on more sophisticated models in the last two Chapters~\ref{chapter:5} and \ref{chapter:6} are built on our library.
%
We detailed three main points of SMPyBandits:
$(i)$ its purpose and qualities,
$(ii)$ its organization,
and $(iii)$ an overview of its usage:
%

@@

$(ii)$
We detailed how SMPyBandits is implementing arms, problems, algorithms, and use these components to implement a simulation loop, with various visualizations being performed after the simulation.
As far as now, SMPyBandits is restricted to the finite-arm case, but it supports a wide range of arms distributions.
Different kinds of models are implemented, from stationary single-player to piece-wise stationary multi-players with different collision models.
One of the main qualities of the library is that it is quite exhaustive, as all the main families of algorithms covering these different models have been implemented, even very recent algorithms from the literature, as we tried our best in following the active research from December $2016$ to June $2019$.
More than $65$ algorithms or variants of algorithms are implemented for the single-player case, $5$ for the expert aggregation problem, about $15$ for the multi-players case, and about $20$ for the piece-wise stationary problem.
All the codebase is fully documented, and the library is using continuous integration to run automated tests on the code after every modification.
%
When comparing algorithms on a problem, the main performance measure is the regret, but the library also computes, stores and visualizes other measures, such as best-arm selection rate, mean cumulated reward, real time or memory costs.

@@

The most exciting thing that could happen to our SMPyBandits library would be to see it gaining popularity!
Its documentation has seen already about $20000$ visits, the project on GitHub had $88$ stars in June $2019$, and based on the features requested and the emails received about it, we counted between $10$ to $20$ researchers in other labs who use or used SMPyBandits.
While this is a good start, we believe that the library is mature and interesting enough to hope to see it being used by more people.
%
We have submitted a summary paper presenting SMPyBandits \cite{SMPyBanditsJMLR} to the MLOSS track of the Journal of Machine Learning Research (JMLR), and we hope that it will be accepted, as it would give more visibility to our work.
%
As a personal note, I would like to continue working on SMPyBandits, and implement the most important requested features, as well as maintain it, and possibly continue to add recent algorithms by following actively the research in this community.
I would also love to be able to teach a graduate or under-graduate course on multi-armed bandits in the next years, and use my library as a support to illustrate the course, as well as practical sessions.

>>>>

\chapter{SMPyBandits: an exhaustive Python library to simulate MAB problems}
% \chapter{SMPyBandits: our Python library to simulate MAB problems}
\label{chapter:3}

@@

\paragraph{References.}
%
The code for SMPyBandits is hosted at \texttt{\href{https://GitHub.com/SMPyBandits/SMPyBandits/}{GitHub.com/SMPyBandits/SMPyBandits}}, its documentation is at \texttt{\href{https://SMPyBandits.GitHub.io/}{SMPyBandits.GitHub.io}}, and all the library is freely publicly released under the MIT open-source License.
This chapter is based on the article \cite{SMPyBanditsJMLR}.

@@

We chose to present here the Python library SMPyBandits in this chapter, rather than in the Appendix, because all the numerical experiments on single-player multi-arm bandits in Chapter~\ref{chapter:25}, and on more sophisticated models in the last two Chapters~\ref{chapter:5} and \ref{chapter:6} are built on SMPyBandits.
%
We detailed three main points of our library:
$(i)$ its purpose and qualities,
$(ii)$ its organization,
and $(iii)$ an overview of its usage:
%

@@

$(ii)$
We detailed how SMPyBandits is implementing arms, problems, algorithms, and use these components to implement a simulation loop, with various visualizations being performed after the simulation.
As far as now, SMPyBandits is restricted to the finite-arm case, but it supports a wide range of arms distributions.
Different kinds of models are implemented, from stationary single-player to piece-wise stationary multi-players with different collision models.
One of the main qualities of the library is that it is quite exhaustive, as all the main families of algorithms covering these different models have been implemented, even very recent algorithms from the literature, as we followed the active research from December $2016$ to June $2019$.
More than $65$ algorithms or variants of algorithms are implemented for the single-player case, $5$ for the expert aggregation problem, about $15$ for the multi-players case, and about $20$ for the piece-wise stationary problem.
All the codebase is fully documented, and the library is using continuous integration to run automated tests on the code after every modification.
%
When comparing algorithms on a problem, the main performance measure is the regret, but the library also computes, stores and visualizes other measures, such as best-arm selection rate, mean cumulated reward, real time or memory costs.

@@

The most exciting thing that could happen to our SMPyBandits library would be to see it gaining popularity!
Its documentation has seen already about $20000$ visits, the project on GitHub had $88$ stars in June $2019$, and based on the features requested and the emails received about it, we counted between $10$ to $20$ researchers in other labs who use or used SMPyBandits.
While this is a good start, we believe that the library is mature and interesting enough to hope to see it being used by more people.
%
We have submitted a summary paper presenting SMPyBandits \cite{SMPyBanditsJMLR} to the MLOSS track of the Journal of Machine Learning Research (JMLR), and we hope that it will be accepted, as it would give more visibility to this work.
%
As a personal note, I would like to continue working on SMPyBandits, and implement the most important requested features, as well as maintain it, and possibly continue to add recent algorithms by following actively the research in this community.
I would also love to be able to teach a graduate or under-graduate course on multi-armed bandits in the next years, and use my library as a support to illustrate the course, as well as practical sessions.

### 2019-07-06 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
J'utilise moins "our ..."

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if the dynamic devices randomly select their channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic for the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communications rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic, more and more fluctuating and unpredictable.

@@

We note that MAB learning has also started to be applied to also optimize the PHY layer parameters, see for instance \cite{KerkoucheAlami18} following our article \cite{Bonnefoi17},
but in this thesis we chose to restrict to the spectrum access problem and thus we only MAB models where the arms correspond to orthogonal frequency bands.

@@

\textbf{Using a bandit algorithm for IoT devices.}
%
Our IoT application is challenging in that there are \emph{multiple} players (the dynamic devices) interacting with the \emph{same} arms (the channels), without any centralized communication (they do not even know the total number of dynamic devices).
%
% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are (unfortunately) \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each device).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

>>>>

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel.
So this first hypothesis is actually not constraining.
We first evaluate the probability of collision if the dynamic devices randomly select their channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic for the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communications rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic, more and more fluctuating and unpredictable.

@@

We note that MAB learning has also started to be applied to also optimize the PHY layer parameters, see for instance \cite{KerkoucheAlami18} which followed our article \cite{Bonnefoi17},
but in this thesis we chose to restrict to the spectrum access problem and thus we only MAB models where the arms correspond to orthogonal frequency bands.

@@

\textbf{Using a bandit algorithm for IoT devices.}
%
The IoT application is challenging in that there are \emph{multiple} players (the dynamic devices) interacting with the \emph{same} arms (the channels), without any centralized communication (they do not even know the total number of dynamic devices).
%
% Considered alone, each dynamic device
We propose algorithms in which each dynamic device ignores all the other one and
implements a learning algorithm to play a bandit game.
%
In each time slot, if it has to communicate (which happens with probability $p$), then it chooses a channel and it receives a reward $1$ if the transmission is successful, $0$ otherwise.
Each device aims at maximizing the sum of the rewards collected during its communication instants, which shall indeed maximize the fraction of successful transmissions. Besides the modified time scale (rewards are no longer collected at every time step), this looks like a usual bandit problem.
However, it cannot be modeled as a stochastic MAB, as the rewards are (unfortunately) \emph{not} \iid: they not only depend on the (stationary, \iid) behavior of the static devices, but also on the behavior of other dynamic devices, that is not stationary (because of learning and random activation of each device).
%
Despite this, we show in the next subsection that running a stochastic bandit algorithm for each device based on its own rewards is surprisingly successful.

### 2019-07-06 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main.tex
J'utilise moins "our ..."

In this section, we present our implementation of MAB algorithms in the model of IoT networks presented in Section~\ref{sub:41:systemModel}.
We first describe the simplified physical layer of this demo, then we present our GNU Radio implementation.

@@

Regarding the physical layer, we consider a QPSK constellation (\emph{Quadrature Phase-Shift Keying}). Moreover, we use simplified packets composed of two parts.
The first part is the \emph{preamble} which is used for the purpose of synchronization (phase correction).
Then, we have the \emph{index} of the user, which is a sequence of QPSK symbols.
For example, this index can be a simple QPSK symbol ($\pm1\pm1j$), or a sequence of QPSK symols if the PoC has to consider more users.
With $\ell$ QPSK symbols, we can indeed fit at most $4^{\ell}/2 = 2^{2\ell-1}$ devices, and not $4^{\ell}$ due to the use of a conjugate index to send back acknowledgements from the base station.
Once the base station receives an uplink packet, it detects this index and transmits an acknowledgement which has the same frame structure, but where the index is the conjugate of the index of the uplink packet ($z \mapsto \overline{z}$, \emph{e.g.}, $1+j \mapsto 1-j$).
Thanks to this index, we can have several devices communicating with the same base station.
%
In turn, the end-device that receives the acknowledgement demodulates it, and checks if the index is the conjugate of its own index.
In this case, the \emph{Ack} was sent for him, and it knows that its packet has been received and decoded correctly by the base station.

@@

\paragraph{Equipment.}
% Boards and versions
We use USRP N210 boards \cite{USRPDocumentation}, from Ettus Research (National Instrument),
with version $4$ of their FPGA system and version $5.1$ of the RBX system.
As illustrated in Figure~\ref{fig:42:our_demo}, our implementation is composed of at least $3$ USRP:
the base station,
a traffic generator which emulates the interfering traffic (made by surrounding static devices),
and at least one dynamic device.
Each dynamic device has its own USRP and its own learning algorithm.
% Cables and laptop
% Osef

@@

We have designed a user interface in order to visualize the results obtained  with our experimental demonstration. This user interface is shown in Figure~\ref{fig:42:UI}.
We can see that it is made of three parts, one for each USRP, as highlighted in \textcolor{darkred}{circled red numbers}.

@@

In Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} below, we show the results averaged on $10$ repetitions using the same conditions.
%
Each experiment has been done on a duration of about half a day,
due to the IoT sporadic transmission mode that we want to respect (like in our model of Section~\ref{sec:4:firstModel}).
However, we make devices generate one message every $5$ seconds, in order to artificially speed up the process and with no loss of generality (as we are using real hardware).
Learning can be useful only when there is a large enough difference between ``good'' and ``bad'' channels,
Each device was learning to access $4$ different non-overlapping channels, that we chose to have occupancy rates of $(\mu_k)_k = [15\%, 10\%, 2\%, 1\%]$.
Note that a maximum occupancy rate of $15\%$ could seem not so high, but indeed it is, because for a pure ALOHA access mode, a naive dynamic device only enjoys about $40\%$ success rate, under such occupancy of the channels (see the ``uniform'' plot in Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS}).
%
The occupancy rate of a channel, which denotes the mean occupancy, is implemented using the traffic generator, to emulate the presence of $S_i$ static devices with emission probability $p$ (that is, $\mu_i = S_i \times p$ here).

@@

Similar results are obtained for overlapping channels, when dynamic devices are learning in the presence of multiple devices, all using the same learning algorithm.
However, our experimental testbed can not run hundreds of intelligent devices.
Empirical results confirm the simulations presented in Section~\ref{sec:4:firstModel} (see Figure~\ref{fig:41:perf_learning}).
Such results are very encouraging, and illustrate well the various strong possibilities of MAB learning applied to IoT networks.

>>>>

In this section, we present the implementation of MAB algorithms in the model of IoT networks presented in Section~\ref{sub:41:systemModel}.
We first describe the simplified physical layer of this demo, then we present our GNU Radio implementation.

@@

Regarding the physical layer, we consider a QPSK constellation (\emph{Quadrature Phase-Shift Keying}, \eg, see references in \cite{Bodinier17}). Moreover, we use simplified packets composed of two parts.
The first part is the \emph{preamble} which is used for the purpose of synchronization (phase correction).
Then, we have the \emph{index} of the user, which is a sequence of QPSK symbols.
For example, this index can be a simple QPSK symbol ($\pm1\pm1j$), or a sequence of QPSK symols if the PoC has to consider more users.
With $\ell$ QPSK symbols, we can indeed fit at most $4^{\ell}/2 = 2^{2\ell-1}$ devices, and not $4^{\ell}$ due to the use of a conjugate index to send back acknowledgements from the base station.
Once the base station receives an uplink packet, it detects this index and transmits an acknowledgement which has the same frame structure, but where the index is the conjugate of the index of the uplink packet ($z \mapsto \overline{z}$, \emph{e.g.}, $1+j \mapsto 1-j$).
Thanks to this index, we can have several devices communicating with the same base station.
%
In turn, the end-device that receives the acknowledgement demodulates it, and checks if the index is the conjugate of its own index.
In this case, the \emph{Ack} was sent for him, and it knows that its packet has been received and decoded correctly by the base station.

@@

\paragraph{Equipment.}
% Boards and versions
We use USRP N210 boards \cite{USRPDocumentation}, from Ettus Research (National Instrument),
with version $4$ of their FPGA system and version $5.1$ of the RBX system.
As illustrated in Figure~\ref{fig:42:our_demo}, the implementation is composed of at least $3$ USRP:
the base station,
a traffic generator which emulates the interfering traffic (made by surrounding static devices),
and at least one dynamic device.
Each dynamic device has its own USRP and its own learning algorithm.
% Cables and laptop
% Osef

@@

We have designed a user interface in order to visualize the results obtained  with the experimental demonstration. This user interface is shown in Figure~\ref{fig:42:UI}.
We can see that it is made of three parts, one for each USRP, as highlighted in \textcolor{darkred}{circled red numbers}.

@@

In Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} below, we show the results averaged on $10$ repetitions using the same conditions.
%
Each experiment has been done on a duration of about half a day,
due to the IoT sporadic transmission mode that we want to respect (like in the model of Section~\ref{sec:4:firstModel}).
However, we make devices generate one message every $5$ seconds, in order to artificially speed up the process and with no loss of generality (as we are using real hardware).
Learning can be useful only when there is a large enough difference between ``good'' and ``bad'' channels,
Each device was learning to access $4$ different non-overlapping channels, that we chose to have occupancy rates of $(\mu_k)_k = [15\%, 10\%, 2\%, 1\%]$.
Note that a maximum occupancy rate of $15\%$ could seem not so high, but indeed it is, because for a pure ALOHA access mode, a naive dynamic device only enjoys about $40\%$ success rate, under such occupancy of the channels (see the ``uniform'' plot in Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS}).
%
The occupancy rate of a channel, which denotes the mean occupancy, is implemented using the traffic generator, to emulate the presence of $S_i$ static devices with emission probability $p$ (that is, $\mu_i = S_i \times p$ here).

@@

Similar results are obtained for overlapping channels, when dynamic devices are learning in the presence of multiple devices, all using the same learning algorithm.
However, our experimental testbed cannot run hundreds of intelligent devices.
Empirical results confirm the simulations presented in Section~\ref{sec:4:firstModel} (see Figure~\ref{fig:41:perf_learning}).
Such results are very encouraging, and illustrate well the various strong possibilities of MAB learning applied to IoT networks.

### 2019-07-06 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
J'utilise moins "our ..."

A device aims to select a channel with the highest probability of successful transmissions, for which it uses a reinforcement learning approach, again formulated as a MAB problem.
Contrarily to what may appear at first sight,
\emph{the goal is not to minimize the number of retransmissions}, but to maximize the probability of successful transmission, considering both the first transmission of a message and the retransmissions of the same message.
Indeed, the objective of each device is to \emph{maximize its battery life by minimizing its \textbf{total} number of transmissions}.
% , where each channel (also called arms) is viewed as a gambling machine (bandit), and each bandit has a \emph{reward}. Then, at every trial, a device chooses a channel that maximizes the sum of the collected rewards. These \emph{rewards} are the \emph{acknowledgment} (\emph{Ack}) signals received after transmitting packets to the gateway. In this way, a transmission is considered successful when an acknowledgment is received, and a learning approach is employed to select the best channel.
%
We address the problem of channel selection taking into account the described Markov model for the retransmissions of end-devices.
% It motivates our present work for which we consider the retransmissions in the analysis of MAB algorithms.
It motivates our present work for which we consider the number of retransmissions, carried out by each device.
% to address several MAB algorithms.
%

@@

We consider IoT devices with a constraint on their QoS, imposing the successful delivery of their messages.
When such an IoT device experiments a collision, it goes in a back-off state to retransmit the same packet on the same or another channel.
If all devices remain in the same channel for retransmissions, it is a well-known result that it could result in a sequence of successive collisions with the same packets' devices that previously collided.
%
Thus, it seems interesting to consider in the decision making policy the possibility for a device to retransmit in a different channel.
One of our motivations to develop new MAB algorithms for our problem is this option of using a different communication channel between the first transmission and the next retransmissions.

@@

We assume to be in the steady state \cite{Norris98}, in our Markov chain model depicted in Figure~\ref{fig:43:Markov_model}, and thus the probabilities no longer depend on the slot number $t$ (\ie, $\forall t, x_t=x$).
Therefore, the probability that this device has a collision at the first transmission is $p_c$, and has the following expression
%
\begin{equation}\label{eq:43:1}
	p_c = 1-\left(1-x\right)^{N-1} \iff x = 1-\left(1-p_c\right)^{\frac{1}{N-1}}.
\end{equation}

@@

In order to assess the proposed approximation, we suppose a unique channel where all the devices follow the same contention Markov process.
We simulate an ALOHA protocol with a maximum number of retransmissions $\mathrm{MaxBackOff}=10$, a maximum back-off interval $m=10$, and a transmission probability $p=10^{-3}$.
%
In Figure~\ref{fig:43:Approximation_m10}, we show the collision probabilities for different number of devices $N$ (from $N=50$ to $N=400$), for both $p_{c}$ and $p_{c1}$.
%
We can verify that our approximation is very precise for lower values $p_{c1} \leq 30 \%$ (\ie, \textcolor{red}{red} and \textcolor{gold}{orange} curves are quite close).
Moreover, a significant gap between $p_{c1}$ and $p_c$,
of up-to $10\%$, can be observed,
which suggests us to resort to MAB algorithms for the channel selection for both the first transmission and next retransmissions.

@@

It is worth to highlight that, if we write \eqref{eq:43:decomppc1} as $p_{c1} = p_c + p_{ca} \left(1-p_c\right)$,
then it is obvious that $p_{c1}$ is always larger than $p_c$ (as $p_{ca} \left(1-p_c\right) > 0$).
But for large values of $p_c$, $p_{ca}\left(1-p_c\right) \simeq 0$ so the gap gets small,
and for small values of $p_c$ the gap is significant.
Moreover, we can verify (\eg, numerically or by differentiating)
that the gap decreases when $p_c$ increases (for fixed $N$ and $m$).
This backups mathematically the observation we made from Figure~\ref{fig:43:Approximation_m10}:
the smaller the $p_c$, the larger is the gap between $p_c$ and $p_{c1}$.
%
We interpret this fact in two different situations:
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then our derivation shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

@@

Although, this approach increases the complexity and storage requirements (now, of order $\mathcal{O}(K^2)$).
For our LPWA networks of interest, such as LoRaWAN, the cost of its implementation is still affordable, since a small number of channels is used.
For instance, for $K=4$ channels,
the memory to store $K+1=5$ algorithms is of the order of the requirements to store one.

@@

We also consider in our experiment the case of an ALOHA protocol using $\mathrm{MaxBackOff}=5$, and $m=10$, a statistic distribution of the devices about $40\%, 30\%, 20\%, 10\%$ for the four channels, and $N=2000$ IoT devices.
The corresponding results are depicted in Figure~\ref{fig:43:mainExperiment2}.
In this case the successful transmission rate is degraded compared with achieved results in Figure~\ref{fig:43:mainExperiment1}.
We can explain this by the fact that we are considering in our network more devices, which increases the collision probability.
It is important to highlight, that the ``Random'' retransmission heuristic shows a poor performance in comparison to the other heuristics, and it can be attributed to the fact that the number of retransmission is increased, and consequently a
learning approach is able to take advantage of it.
Furthermore, the ``\UCB'', ``$K$ \UCB'' and ``Delayed \UCB'' heuristics behave similarly to ``Only \UCB'', after a similar convergence time.

>>>>

A device aims to select a channel with the highest probability of successful transmissions, for which it uses a reinforcement learning approach, again formulated as a MAB problem.
Contrarily to what may appear at first sight,
\emph{the goal is not to minimize the number of retransmissions}, but to maximize the probability of successful transmission, considering both the first transmission of a message and the retransmissions of the same message.
Indeed, the objective of each device is to \emph{maximize its battery life by minimizing its \textbf{total} number of transmissions}.
% , where each channel (also called arms) is viewed as a gambling machine (bandit), and each bandit has a \emph{reward}. Then, at every trial, a device chooses a channel that maximizes the sum of the collected rewards. These \emph{rewards} are the \emph{acknowledgment} (\emph{Ack}) signals received after transmitting packets to the gateway. In this way, a transmission is considered successful when an acknowledgment is received, and a learning approach is employed to select the best channel.
%
We address the problem of channel selection taking into account the described Markov model for the retransmissions of end-devices.
% It motivates our present work for which we consider the retransmissions in the analysis of MAB algorithms.
It motivates this section for which we consider the number of retransmissions, carried out by each device.
% to address several MAB algorithms.
%

@@

We consider IoT devices with a constraint on their QoS, imposing the successful delivery of their messages.
When such an IoT device experiments a collision, it goes in a back-off state to retransmit the same packet on the same or another channel.
If all devices remain in the same channel for retransmissions, it is a well-known result that it could result in a sequence of successive collisions with the same packets' devices that previously collided.
%
Thus, it seems interesting to consider in the decision making policy the possibility for a device to retransmit in a different channel.
One of our motivations to develop new MAB algorithms for the considered problem is this option of using a different communication channel between the first transmission and the next retransmissions.

@@

We assume to be in the steady state \cite{Norris98}, in the Markov chain model depicted in Figure~\ref{fig:43:Markov_model}, and thus the probabilities no longer depend on the slot number $t$ (\ie, $\forall t, x_t=x$).
Therefore, the probability that this device has a collision at the first transmission is $p_c$, and has the following expression
%
\begin{equation}\label{eq:43:1}
	p_c = 1-\left(1-x\right)^{N-1} \iff x = 1-\left(1-p_c\right)^{\frac{1}{N-1}}.
\end{equation}

@@

In order to assess the proposed approximation, we suppose a unique channel where all the devices follow the same contention Markov process.
We simulate an ALOHA protocol with a maximum number of retransmissions $\mathrm{MaxBackOff}=10$, a maximum back-off interval $m=10$, and a transmission probability $p=10^{-3}$.
%
In Figure~\ref{fig:43:Approximation_m10}, we show the collision probabilities for different number of devices $N$ (from $N=50$ to $N=400$), for both $p_{c}$ and $p_{c1}$.
%
We can verify that the obtained approximation is very precise for lower values $p_{c1} \leq 30 \%$ (\ie, \textcolor{red}{red} and \textcolor{gold}{orange} curves are quite close).
Moreover, a significant gap between $p_{c1}$ and $p_c$,
of up-to $10\%$, can be observed,
which suggests us to resort to MAB algorithms for the channel selection for both the first transmission and next retransmissions.

@@

It is worth to highlight that, if we write \eqref{eq:43:decomppc1} as $p_{c1} = p_c + p_{ca} \left(1-p_c\right)$,
then it is obvious that $p_{c1}$ is always larger than $p_c$ (as $p_{ca} \left(1-p_c\right) > 0$).
But for large values of $p_c$, $p_{ca}\left(1-p_c\right) \simeq 0$ so the gap gets small,
and for small values of $p_c$ the gap is significant.
Moreover, we can verify (\eg, numerically or by differentiating)
that the gap decreases when $p_c$ increases (for fixed $N$ and $m$).
This backups mathematically the observation we made from Figure~\ref{fig:43:Approximation_m10}:
the smaller the $p_c$, the larger is the gap between $p_c$ and $p_{c1}$.
%
We interpret this fact in two different situations:
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then the derivation above shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

@@

Although, this approach increases the complexity and storage requirements (now, of order $\mathcal{O}(K^2)$).
For the LPWA networks of interest, such as LoRaWAN, the cost of its implementation is still affordable, since a small number of channels is used.
For instance, for $K=4$ channels,
the memory to store $K+1=5$ algorithms is of the order of the requirements to store one.

@@

We also consider in these experiment the case of an ALOHA protocol using $\mathrm{MaxBackOff}=5$, and $m=10$, a statistic distribution of the devices about $40\%, 30\%, 20\%, 10\%$ for the four channels, and $N=2000$ IoT devices.
The corresponding results are depicted in Figure~\ref{fig:43:mainExperiment2}.
In this case the successful transmission rate is degraded compared with achieved results in Figure~\ref{fig:43:mainExperiment1}.
We can explain this by the fact that we are considering in this network an increased number of devices, which increases the collision probability.
It is important to highlight, that the ``Random'' retransmission heuristic shows a poor performance in comparison to the other heuristics, and it can be attributed to the fact that the number of retransmission is increased, and consequently a
learning approach is able to take advantage of it.
Furthermore, the ``\UCB'', ``$K$ \UCB'' and ``Delayed \UCB'' heuristics behave similarly to ``Only \UCB'', after a similar convergence time.

### 2019-07-06 2-Chapters/4-Chapter/chapter4.tex
Small modifs
J'utilise moins "our ..."

\newpage  % WARNING ?
% ----------------------------------------------------------------------------
\section{Test-bed implementation of our model for real-world validation}
\label{sec:4:gnuradio}
% ----------------------------------------------------------------------

@@

It is also quite surprising that stochastic MAB algorithms can be of any use in such non-stationary applications.
Unfortunately, analyzing our model with thousands of independent devices, all communicating and learning in their own (random) time scales, turned out to be of extreme difficulty.
That is why we focus on two different directions of theoretical analysis in the next chapters of this thesis.
% %
% We can also highlight that the two models we studied are complementary:
% the first model (without retransmission of packets), is interesting for its simplicity, and focused on improving the battery life of IoT devices,
% % , at the cost of a lower Quality of Service (QoS),
% %
% while the second model is more interesting if the main target is an improvement of the Quality of Service (QoS).
% % , rather than a longer battery life.

>>>>

\newpage  % WARNING ?
% ----------------------------------------------------------------------------
% \section{Test-bed implementation of the first model for real-world validation}
\section{Proof-of-concept of our model for real-world validation}
\label{sec:4:gnuradio}
% ----------------------------------------------------------------------

@@

It is also quite surprising that stochastic MAB algorithms can be of any use in such non-stationary applications.
Unfortunately, analyzing the considered model with thousands of independent devices, all communicating and learning in their own (random) time scales, turned out to be of extreme difficulty.
That is why we focus on two different simplifications of this model in the next chapters, for which we are able to provide clean theoretical analyses.
% %
% We can also highlight that the two models we studied are complementary:
% the first model (without retransmission of packets), is interesting for its simplicity, and focused on improving the battery life of IoT devices,
% % , at the cost of a lower Quality of Service (QoS),
% %
% while the second model is more interesting if the main target is an improvement of the Quality of Service (QoS).
% % , rather than a longer battery life.
